Cubical Polarized Types

### 2014-05-20 ring.tex
fixed values I think and started to think about combinators

\documentclass[authoryear,preprint]{sigplanconf}
\usepackage{amsmath}
\usepackage{listings} 
\usepackage{stmaryrd}
\usepackage{latexsym}
\usepackage{amssymb}
\usepackage{xcolor}
\usepackage{courier}
\usepackage{thmtools}
\usepackage{bbold}
\usepackage{tikz}
\usepackage{proof}

\newcommand{\negp}[1]{\textit{neg}(#1)}
\newcommand{\inl}[1]{\textsf{inl}~#1}
\newcommand{\inr}[1]{\textsf{inr}~#1}
\newcommand{\lolli}{\multimap} 
\newcommand{\cubt}{\mathbb{T}}
\newcommand{\den}[1]{\llbracket #1 \rrbracket}
\newcommand{\denc}[1]{\llbracket #1 \rrbracket^c}
\newcommand{\nodet}[2]{\fcolorbox{black}{white}{$#1$}\fcolorbox{black}{gray!20}{$#2$}}

@@

We lose the multiplicative structure. No evident way to define the
multiplication functor. Perhaps there is a clever way. But this turns out to
be a well-known open problem. Review the problem and tell the story from the
algebraic topology perspective.

Regular first-order types are viewed as 0-dimensional cubes. Generalize to
$n$-dimensional cubes.

@@

We first define the syntax and semantics of types and then discuss type
isomorphisms.

@@

We use $t$ for regular first-order types which include the empty type, the
unit type, and conventional sum and product types.  We use $\tau$ for cubical
types: these have the same formal structure as regular types but also include
\emph{negative} types. The syntax of the types is summarized below:
\[\begin{array}{rcl}
t &::=& 0 \alt 1 \alt t_1 + t_2 \alt t_1 * t_2 \\
\tau &::=& t \alt 
      \alt \tau_1 + \tau_2
      \alt \tau_1 * \tau_2
      \alt - \tau
\end{array}\]
We use $\tau_1 - \tau_2$ to abbreviate $\tau_1 + (- \tau_2)$ and more
interestingly $\tau_1 \lolli \tau_2$ to abbreviate $(- \tau_1) + \tau_2$. At
the level of cubical types, we can write types such as
$(t_1-t_2)-(t_3-t_4*t_5)$ which mix negative types with the usual type
constructors.

It is easy to find a model for regular types: finite sets are such a
model. In more detail:
\[\begin{array}{rcl}
\den{0} &=& \emptyset \\
\den{1} &=& \{ \bullet \} \\
\den{t_1 + t_2} &=& \den{t_1} \uplus \den{t_2} \\
\den{t_1 * t_2} &=& \den{t_1} \times \den{t_2} 
\end{array}\]
The type 0 maps to the empty set. The type 1 maps to a singleton set. The sum
of types maps to disjoint union of sets. The product of types maps to the
cartesian product of sets. We usually use $S$ for the denotation of a
conventional first-order type.

For cubical types, their denotations, usually written as $\cubt$, are
$n$-dimensional cubical sets with polarities. An $n$-dimensional cubical set
$\nodet{\cubt_1}{\cubt_2}$ consists of a positive subspace $\cubt_1$ and a
negative subspace $\cubt_2$ (shaded in gray). Each of the subspaces are
cubical set of a lower dimension; the 0-dimensional case reduces to the
denotation of conventional first-order types. As we explain below, cubical
sets are closed under sums, differences, products.

Examples of 0-dimensional cubical sets include all the finite sets denoting a
regular first-order type. A 1-dimensional cubical set looks like
$\nodet{S_1}{S_2}$ which intuitively corresponds to the difference $t_1 -
t_2$ of the two types whose denotations are $S_1$ and $S_2$ respectively. A
2-dimensional cubical type looks like
$\nodet{(\nodet{S_1}{S_2})}{(\nodet{S_3}{S_4})}$ which intuitively
corresponds to the iterated difference of the appropriate types
$(t_1-t_2)-(t_3-t_4)$ where the product of the successive signs from the
outermost box encodes the sign.

We now give the denotation of cubical types:
\[\begin{array}{rcl}
\denc{t} &=& \den{t} \\
\denc{\tau_1 + \tau_2} &=& \denc{\tau_1} \oplus \denc{\tau_2} \\
\denc{\tau_1 * \tau_2} &=& \denc{\tau_1} \otimes \denc{\tau_2} \\
\denc{- \tau} &=& \ominus \denc{\tau}
\end{array}\]
where the operations $\oplus$, $\otimes$, and $\ominus$ are defined below.
It is possible for cubical sets not be balanced in general but it is always
possible to embed an $n$-dimensional set into a larger dimension using the
fact (to be justified later) that $t$ is equivalent to $t-0$ and to $(t-0)-0$
and so on, i.e.,
\[
S = \nodet{S}{\phantom{S}} = 
  \nodet{\nodet{S}{\phantom{S}}}
        {\phantom{{\nodet{S}{\phantom{S}}}}} = \ldots \\
\]
In the recursive definitions below, we assume that the cubical sets have been
extended to be balanced as needed:
\[\begin{array}{rcl}
\ominus~S &=& \nodet{\phantom{S}}{S} \\
\ominus~\nodet{\cubt_1}{\cubt_2} &=& \nodet{\ominus~\cubt_2}{\ominus~\cubt_1} \\
\\
S_1 \oplus S_2 &=& S_1 \uplus S_2 \\
(\nodet{\cubt_1}{\cubt_2}) \oplus (\nodet{\cubt_3}{\cubt_4}) &=& 
  \nodet{\cubt_1 \oplus \cubt_3}{\cubt_2 \oplus \cubt_4} \\
\\
S_1 \otimes S_2 &=& S_1 \times S_2 \\
S \otimes (\nodet{\cubt_1}{\cubt_2}) &=& 
  \nodet{S \otimes \cubt_1}{S \otimes \cubt_2} \\
(\nodet{\cubt_1}{\cubt_2}) \otimes \cubt &=& 
  \nodet{\cubt_1 \otimes \cubt}{\cubt_2 \otimes \cubt}
\end{array}\]

@@

\noindent The definition of negation is natural: it recursively swaps the
positive and negative subspaces. We note that the general definitions of sum
and product reduce to the conventional sums and products for 0-dimensional
types. Also the sum of 1-dimensional types reduces to the sum used in the
\textbf{Int} construction and that addition and negation do not increase the
dimensions. The interesting new development is the product definition. The
example in Fig.~\ref{mult} illustrates a general pattern: the product of an
$n$-dimensional type with an $m$-dimensional type gives an
$(n+m)$-dimensional type. For low dimensions such as the ones used in the
example above, it is possible to visualize what is happening in coordinate
space (see Fig.~\ref{multv}).

\begin{figure*}
\begin{tikzpicture}
\node[left] at (0,0) {$S_1^+$};
\draw[fill] (0,0) circle [radius=0.05];
\node[right] at (0.6,0) {$S_2^-$};
\draw[fill] (0.6,0) circle [radius=0.05];
\draw[-] (0,0) -- (0.6,0);
\node at (1.6,0) {$\otimes$};

%%
\node[below left] at (2.5,-0.5) {$S_3^+$};
\draw[fill] (2.5,-0.5) circle [radius=0.05];
\node[below right] at (3.5,-0.5) {$S_4^-$};
\draw[fill] (3.5,-0.5) circle [radius=0.05];
\draw[-] (2.5,-0.5) -- (3.5,-0.5);
\draw[-] (2.5,-0.5) -- (2.5,0.5);
\node[above left] at (2.5,0.5) {$S_5^-$};
\draw[fill] (2.5,0.5) circle [radius=0.05];
\node[above right] at (3.5,0.5) {$S_6^+$};
\draw[fill] (3.5,0.5) circle [radius=0.05];
\draw[-] (2.5,0.5) -- (3.5,0.5);
\draw[-] (3.5,-0.5) -- (3.5,0.5);
%% 
\node at (5,0) {$=$};
%% 
\node[left] at (7.5,0.75) {$(S_2 \times S_3)^-$};
\draw[fill] (7.5,0.75) circle [radius=0.05];
\node[right] at (9.5,0.75) {$(S_2 \times S_4)^+$};
\draw[fill] (9.5,0.75) circle [radius=0.05];
\node[above right] at (10.2,1.2) {$(S_2 \times S_6)^-$};
\draw[fill] (10.2,1.2) circle [radius=0.05];
\node[above left] at (8.2,1.2) {$(S_2 \times S_5)^+$};
\draw[fill] (8.2,1.2) circle [radius=0.05];
%%
\node[left] at (7.5,-0.75) {$(S_1 \times S_3)^+$};
\draw[fill] (7.5,-0.75) circle [radius=0.05];
\node[right] at (9.5,-0.75) {$(S_1 \times S_4)^-$};
\draw[fill] (9.5,-0.75) circle [radius=0.05];
\node[above right] at (10.2,-0.3) {$(S_1 \times S_6)^+$};
\draw[fill] (10.2,-0.3) circle [radius=0.05];
\node[above] at (8.2,-0.3) {$(S_1 \times S_5)^-$};
\draw[fill] (8.2,-0.3) circle [radius=0.05];
%%
\draw[-] (7.5,0.75) -- (9.5,0.75);
\draw[-] (9.5,0.75) -- (10.2,1.2);
\draw[-] (10.2,1.2) -- (8.2,1.2);
\draw[-] (8.2,1.2) -- (7.5,0.75);
%%
\draw[-] (7.5,-0.75) -- (9.5,-0.75);
\draw[-] (9.5,-0.75) -- (10.2,-0.3);
\draw[-,dashed] (10.2,-0.3) -- (8.2,-0.3);
\draw[-,dashed] (8.2,-0.3) -- (7.5,-0.75);
%%
\draw[-] (7.5,0.75) -- (7.5,-0.75);
\draw[-] (9.5,0.75) -- (9.5,-0.75);
\draw[-] (10.2,1.2) -- (10.2,-0.3);
\draw[-,dashed] (8.2,1.2) -- (8.2,-0.3);
\end{tikzpicture}
\caption{\label{multv}Visualization of multiplication of two cubical sets.}
\end{figure*}

@@

Now that the type structure is defined, we turn our attention to the notion
of values. For conventional first-order types, the values are also
conventional:
\[
\infer{() : 1}{} 
\qquad
\infer{\inl{v} : t_1 + t_2}{v : t_1}
\qquad
\infer{\inr{v} : t_1 + t_2}{v : t_2}
\qquad
\infer{(v_1,v_2) : t_1 * t_2}{v_1 : t_1 & v_2 : t_2}
\]
For cubical types, the values are more involved. Intuitively, a value of the
$n$-dimensional type $\tau$ is an element of one of the sets located in a
corner of the $n$-dimensional cube denoted by $\tau$. Thus to specify a
value, we must specify one of the corners of the cube which can easily be
done using a sequence $p$ of $+$ and $-$ polarities indicating how to
navigate the cube in each successive dimension starting from a fixed origin
to reach the desired corner. We denote such values as $v^{p}$ and use
$\epsilon$ for the empty sequence of polarities. One can think of the
sequence of signs as denoting a path in the binary tree representation of
cubical types. Formally:
\[\begin{array}{c}
\infer[\textit{base}]{v^{\epsilon} : t}{v : t} \\
\\
\infer[\textit{left}]{(\inl{v^p})^{p} : \tau_1 + \tau_2}{v^{p} : \tau_1}
\qquad
\infer[\textit{right}]{(\inr{v^p})^{p} : \tau_1 + \tau_2}{v^{p} : \tau_2} \\
\\
\infer[\textit{prod}]{(v_1^{p_1},v_2^{p_2})^{p_1 \bullet p_2} : \tau_1 * \tau_2}
      {v_1^{p_1} : \tau_1 & v_2^{p_2} : \tau_2} \\
\\
\infer[\textit{neg}]{v^{\negp{p}} : - \tau}{v^p : \tau} 
\end{array}\]
The rule \textit{base} states that a conventional first-order value can be
viewed as trivial 0-dimensional cubical value. The rules \textit{left} and
\textit{right} reflect the fact that sums do not increase the dimension and
works pointwise. The rule \textit{prod} is the most involved one: it
increases the dimension as calculated below:
\[\begin{array}{rcl}
\epsilon \bullet p &=& p \\
(+p_1) \bullet p_2 &=& + (p_1 \bullet p_2) \\
(-p_1) \bullet p_2 &=& - (p_1 \bullet p_2) 
\end{array}\]
The rule \textit{neg} shows that a negative value $v$ is located at the
``opposite'' corner of $v$ which is calculated as follows:
\[\begin{array}{rcl}
\negp{\epsilon} &=& -\epsilon \\
\negp{+p} &=& -\negp{p} \\
\negp{-p} &=& +\negp{p}
\end{array}\]

>>>>

\documentclass[authoryear,preprint]{sigplanconf}
\usepackage{amsmath}
\usepackage{listings} 
\usepackage{stmaryrd}
\usepackage{latexsym}
\usepackage{amssymb}
\usepackage{xcolor}
\usepackage{courier}
\usepackage{thmtools}
\usepackage{bbold}
\usepackage{tikz}
\usepackage{proof}
\usepackage{graphicx}

\newcommand{\Rule}[4]{
\makebox{{\rm #1}
$\displaystyle
\frac{\begin{array}{l}#2\\\end{array}}
{\begin{array}{l}#3\\\end{array}}$
 #4}}
\newcommand{\proves}{\vdash}
\newcommand{\symc}[1]{\mathit{sym}~#1}
\newcommand{\jdg}[3]{#2 \proves_{#1} #3}
\newcommand{\adjoint}[1]{#1^{\dagger}}
\newcommand{\iso}{\leftrightarrow}
\newcommand{\identlp}{\mathit{identl}_+}
\newcommand{\identrp}{\mathit{identr}_+}
\newcommand{\swapp}{\mathit{swap}_+}
\newcommand{\assoclp}{\mathit{assocl}_+}
\newcommand{\assocrp}{\mathit{assocr}_+}
\newcommand{\identlt}{\mathit{identl}_*}
\newcommand{\identrt}{\mathit{identr}_*}
\newcommand{\swapt}{\mathit{swap}_*}
\newcommand{\assoclt}{\mathit{assocl}_*}
\newcommand{\assocrt}{\mathit{assocr}_*}
\newcommand{\distz}{\mathit{dist}_0}
\newcommand{\factorz}{\mathit{factor}_0}
\newcommand{\dist}{\mathit{dist}}
\newcommand{\factor}{\mathit{factor}}
\newcommand{\idc}{\mathit{id}}
\newcommand{\plus}{\raisebox{.4\height}{\scalebox{.6}{+}}}
\newcommand{\minus}{\raisebox{.4\height}{\scalebox{.8}{-}}}
\newcommand{\mm}{\texttt{-}}
\newcommand{\pp}{\texttt{+}}
\newcommand{\negp}[1]{\textit{neg}(#1)}
\newcommand{\inl}[1]{\textsf{inl}~#1}
\newcommand{\inr}[1]{\textsf{inr}~#1}
\newcommand{\lolli}{\multimap} 
\newcommand{\cubt}{\mathbb{T}}
\newcommand{\den}[1]{\llbracket #1 \rrbracket}
\newcommand{\denc}[1]{\llbracket #1 \rrbracket^c}
\newcommand{\nodet}[2]{\fcolorbox{black}{white}{$#1$}\fcolorbox{black}{gray!20}{$#2$}}

@@

We lose the multiplicative structure. No evident way to define the
multiplication functor. Perhaps there is a clever way. But this turns out to
be a well-known open problem. Review the problem and tell the story from the
algebraic topology perspective. Regular first-order types are viewed as
$0$-dimensional cubes. Generalize to $n$-dimensional cubes.

@@

We first define the syntax a semantic model of types and then discuss type
isomorphisms.

@@

We use $t$ for regular first-order types which include the empty type, the
unit type, and conventional sum and product types. We call these types
$0$-dimensional. We use $\tau$ for cubical types of dimension 1 or higher:
these have the same syntactic structure as regular types except that they
also include \emph{negative} types. The syntax of the types is summarized
below:
\[\begin{array}{lrcl}
\textit{types of dimension $0$} & 
  t &::=& 0 \alt 1 \alt t_1 + t_2 \alt t_1 * t_2 \\
\textit{types of dimensions $\geq 1$} & 
  \tau &::=& t \alt \alt \tau_1 + \tau_2 \alt \tau_1 * \tau_2 \alt - \tau
\end{array}\]
We use $\tau_1 - \tau_2$ to abbreviate $\tau_1 + (- \tau_2)$ and more
interestingly $\tau_1 \lolli \tau_2$ to abbreviate $(- \tau_1) + \tau_2$.

It is easy to find a model for $0$-dimensional types: finite sets are such a
model. In more detail:
\[\begin{array}{rcl}
\den{0} &=& \emptyset \\
\den{1} &=& \{ \bullet \} \\
\den{t_1 + t_2} &=& \den{t_1} \uplus \den{t_2} \\
\den{t_1 * t_2} &=& \den{t_1} \times \den{t_2} 
\end{array}\]
The type 0 maps to the empty set. The type 1 maps to a singleton set. The sum
of types maps to disjoint union of sets. The product of types maps to the
cartesian product of sets. We usually use $S$ for the denotation of these
conventional first-order types. Note that our (simple) semantics of types
identifies several structurally different types such as $(1+(1+1))$ and
$((1+1)+1)$. In some sense, this is innocent as the types are
isomorphic. However, in the operational semantics discussed in the next
section, we make the computational content of such type isomorphisms
explicit.

For cubical types of dimension 1 or higher, their denotations, usually
written as $\cubt$, are $n$-dimensional ``polarized cubes.'' Such an
$n$-dimensional cube $\nodet{\cubt_1}{\cubt_2}$ consists of a positive
subspace $\cubt_1$ and a negative subspace $\cubt_2$ (shaded in gray). Each
of the subspaces are cubes of a lower dimension; the $0$-dimensional case
reduces to the denotation of conventional first-order types. As we explain
below, cubical sets are closed under sums, differences, products.

A 1-dimensional cube looks like $\nodet{S_1}{S_2}$ which intuitively
corresponds to the difference $t_1 - t_2$ of the two types whose denotations
are $S_1$ and $S_2$ respectively. A 2-dimensional cube looks like
$\nodet{(\nodet{S_1}{S_2})}{(\nodet{S_3}{S_4})}$ which intuitively
corresponds to the iterated difference of the appropriate types
$(t_1-t_2)-(t_3-t_4)$ where the product of the successive signs from the
outermost box encodes the sign.

We now give the full denotation of cubical types:
\[\begin{array}{rcl}
\denc{t} &=& \nodet{\den{t}}{\phantom{\den{t}}} \\
\denc{\tau_1 + \tau_2} &=& \denc{\tau_1} \oplus \denc{\tau_2} \\
\denc{\tau_1 * \tau_2} &=& \denc{\tau_1} \otimes \denc{\tau_2} \\
\denc{- \tau} &=& \ominus \denc{\tau}
\end{array}\]
where the operations $\oplus$, $\otimes$, and $\ominus$ are defined below.
It is possible, in general, for the binary trees representing the cubes not
be balanced. It is however always possible to embed an $n$-dimensional cube
into a cube of a larger dimension using the fact (to be justified using a
$\Pi$-combinator that witnesses this isomorphism in the next section) that
$t$ is equivalent to $t-0$ and to $(t-0)-0$ and so on, i.e.,
\[
S = \nodet{S}{\phantom{S}} = 
  \nodet{\nodet{S}{\phantom{S}}}
        {\phantom{{\nodet{S}{\phantom{S}}}}} = \ldots \\
\]
In the recursive definitions below, we assume that the cubical sets have been
extended to be balanced as needed:
\[\begin{array}{rcl}
\ominus~S &=& \nodet{\phantom{S}}{S} \\
\ominus~\nodet{\cubt_1}{\cubt_2} &=& \nodet{\ominus~\cubt_2}{\ominus~\cubt_1} \\
\\
S_1 \oplus S_2 &=& S_1 \uplus S_2 \\
(\nodet{\cubt_1}{\cubt_2}) \oplus (\nodet{\cubt_3}{\cubt_4}) &=& 
  \nodet{\cubt_1 \oplus \cubt_3}{\cubt_2 \oplus \cubt_4} \\
\\
S_1 \otimes S_2 &=& S_1 \times S_2 \\
S \otimes (\nodet{\cubt_1}{\cubt_2}) &=& 
  \nodet{S \otimes \cubt_1}{S \otimes \cubt_2} \\
(\nodet{\cubt_1}{\cubt_2}) \otimes \cubt &=& 
  \nodet{\cubt_1 \otimes \cubt}{\cubt_2 \otimes \cubt}
\end{array}\]

@@

\noindent The definition of negation is natural: it recursively swaps the
positive and negative subspaces. We note that the general definitions of sums
and products reduce to the conventional sums and products for $0$-dimensional
types. Also the sum of 1-dimensional types reduces to the sum used in the
\textbf{Int} construction and that addition and negation do not increase the
dimensions. The interesting new development is the definition of
products. The example in Fig.~\ref{mult} illustrates a general pattern: the
product of an $n$-dimensional type with an $m$-dimensional type gives an
$(n+m)$-dimensional type. For low dimensions such as the ones used in the
example above, it is possible to visualize what is happening in coordinate
space (see Fig.~\ref{multv}).

\begin{figure*}
\begin{tikzpicture}
\node[above] at (0,0) {\pp};
\node[left] at (0,0) {$S_1$};
\draw[fill] (0,0) circle [radius=0.05];
\node[above] at (0.6,0) {\mm};
\node[right] at (0.6,0) {$S_2$};
\draw[fill] (0.6,0) circle [radius=0.05];
\draw[-] (0,0) -- (0.6,0);
\node at (1.6,0) {$\otimes$};

%%
\node[left] at (2.5,-0.5) {\pp\pp};
\node[below] at (2.5,-0.5) {$S_3$};
\draw[fill] (2.5,-0.5) circle [radius=0.05];
\node[right] at (3.5,-0.5) {\pp\mm};
\node[below] at (3.5,-0.5) {$S_4$};
\draw[fill] (3.5,-0.5) circle [radius=0.05];
\draw[-] (2.5,-0.5) -- (3.5,-0.5);
\draw[-] (2.5,-0.5) -- (2.5,0.5);
\node[left] at (2.5,0.5) {\mm\pp};
\node[above] at (2.5,0.5) {$S_5$};
\draw[fill] (2.5,0.5) circle [radius=0.05];
\node[right] at (3.5,0.5) {\mm\mm};
\node[above] at (3.5,0.5) {$S_6$};
\draw[fill] (3.5,0.5) circle [radius=0.05];
\draw[-] (2.5,0.5) -- (3.5,0.5);
\draw[-] (3.5,-0.5) -- (3.5,0.5);
%% 
\node at (5,0) {$=$};
%% 
\node[left] at (7.5,0.75) {$(S_2 \times S_3)$\,\mm\pp\pp};
\draw[fill] (7.5,0.75) circle [radius=0.05];
\node[right] at (9.5,0.75) {$(S_2 \times S_4)$\,\mm\pp\mm};
\draw[fill] (9.5,0.75) circle [radius=0.05];
\node[above right] at (10.2,1.2) {$(S_2 \times S_6)$\,\mm\mm\mm};
\draw[fill] (10.2,1.2) circle [radius=0.05];
\node[above left] at (8.2,1.2) {$(S_2 \times S_5)$\,\mm\mm\pp};
\draw[fill] (8.2,1.2) circle [radius=0.05];
%%
\node[left] at (7.5,-0.75) {$(S_1 \times S_3)$\,\pp\pp\pp};
\draw[fill] (7.5,-0.75) circle [radius=0.05];
\node[right] at (9.5,-0.75) {$(S_1 \times S_4)$\,\pp\pp\mm};
\draw[fill] (9.5,-0.75) circle [radius=0.05];
\node[above right] at (10.2,-0.3) {$(S_1 \times S_6)$\,\pp\mm\mm};
\draw[fill] (10.2,-0.3) circle [radius=0.05];
\node[left] at (8.2,-0.3) {$(S_1 \times S_5)$\,\pp\mm\pp};
\draw[fill] (8.2,-0.3) circle [radius=0.05];
%%
\draw[-] (7.5,0.75) -- (9.5,0.75);
\draw[-] (9.5,0.75) -- (10.2,1.2);
\draw[-] (10.2,1.2) -- (8.2,1.2);
\draw[-] (8.2,1.2) -- (7.5,0.75);
%%
\draw[-] (7.5,-0.75) -- (9.5,-0.75);
\draw[-] (9.5,-0.75) -- (10.2,-0.3);
\draw[-,dashed] (10.2,-0.3) -- (8.2,-0.3);
\draw[-,dashed] (8.2,-0.3) -- (7.5,-0.75);
%%
\draw[-] (7.5,0.75) -- (7.5,-0.75);
\draw[-] (9.5,0.75) -- (9.5,-0.75);
\draw[-] (10.2,1.2) -- (10.2,-0.3);
\draw[-,dashed] (8.2,1.2) -- (8.2,-0.3);
\end{tikzpicture}
\caption{\label{multv}Visualization of multiplication of two cubical sets.}
\end{figure*}

@@

We first define values then combinators that manipulate the values to witness
the type isomorphisms.

%%%%%%%%%%%%%%%%%%
\subsection{Values}

Now that the type structure is defined, we turn our attention to the notion
of values. For conventional first-order types, the values are also
conventional:
\[
\infer{() : 1}{} 
\qquad
\infer{\inl{v} : t_1 + t_2}{v : t_1}
\qquad
\infer{\inr{v} : t_1 + t_2}{v : t_2}
\qquad
\infer{(v_1,v_2) : t_1 * t_2}{v_1 : t_1 & v_2 : t_2}
\]
For cubical types, the values are more involved. Intuitively, a value of the
$n$-dimensional type $\tau$ is an element of one of the sets located in a
corner of the $n$-dimensional cube denoted by $\tau$. Thus to specify a
value, we must specify one of the corners of the cube which can easily be
done using a sequence $p$ of $+$ and $-$ polarities indicating how to
navigate the cube in each successive dimension starting from a fixed origin
to reach the desired corner. We denote such values as $v^{p}$ and use
$\epsilon$ for the empty sequence of polarities. One can think of the
sequence of signs as denoting a path in the binary tree representation of
cubical types. Formally:
\[\begin{array}{c}
\infer[\textit{base}]{v^{+} : t}{v : t} \\
\\
\infer[\textit{left}]{(\inl{v})^{p} : \tau_1 + \tau_2}{v^{p} : \tau_1}
\qquad
\infer[\textit{right}]{(\inr{v})^{p} : \tau_1 + \tau_2}{v^{p} : \tau_2} \\
\\
\infer[\textit{prod}]{(v_1^{p_1},v_2^{p_2})^{p_1 \cdot p_2} : \tau_1 * \tau_2}
      {v_1^{p_1} : \tau_1 & v_2^{p_2} : \tau_2} \\
\\
\infer[\textit{neg}]{v^{\negp{p}} : - \tau}{v^p : \tau} 
\end{array}\]
The rule \textit{base} states that a conventional first-order value can be
viewed as trivial $0$-dimensional cubical value. The rules \textit{left} and
\textit{right} reflect the fact that sums do not increase the dimension and
works pointwise. The rule \textit{prod} is the most involved one: it
increases the dimension by \emph{concatenating} the two dimensions in its
arguments. For example, if we pair $v_1^{\pp}$ and $v_2^{\mm\pp}$ we get
$(v_1,v_2)^{\pp\mm\pp}$. The rule \textit{neg} shows that a negative value
$v$ is located at the ``opposite'' corner of $v$ which is calculated as
follows:
\[\begin{array}{rcl}
\negp{\epsilon} &=& \epsilon \\
\negp{+p} &=& -\negp{p} \\
\negp{-p} &=& +\negp{p}
\end{array}\]

%%%%%%%%%%%%%%%%%%
\subsection{Combinators}

The terms of $\Pi$ witness type isomorphisms of the form $b \iso b$. They
consist of base isomorphisms, as defined below, and their composition.
\[\begin{array}{rrcll}
\identlp :&  0 + b & \iso & b &: \identrp \\
\swapp :&  b_1 + b_2 & \iso & b_2 + b_1 &: \swapp \\
\assoclp :&  b_1 + (b_2 + b_3) & \iso & (b_1 + b_2) + b_3 &: \assocrp \\
\identlt :&  1 * b & \iso & b &: \identrt \\
\swapt :&  b_1 * b_2 & \iso & b_2 * b_1 &: \swapt \\
\assoclt :&  b_1 * (b_2 * b_3) & \iso & (b_1 * b_2) * b_3 &: \assocrt \\
\distz :&~ 0 * b & \iso & 0 &: \factorz \\
\dist :&~ (b_1 + b_2) * b_3 & \iso & (b_1 * b_3) + (b_2 * b_3)~ &: \factor
\end{array}\]
Each line of the above table introduces a pair of dual
constants\footnote{where $\swapp$ and $\swapt$ are self-dual.} that witness
the type isomorphism in the middle.  These are the base (non-reducible) terms
of the second, principal level of $\Pi$. Note how the above has two readings:
first as a set of typing relations for a set of constants. Second, if these
axioms are seen as universally quantified, orientable statements, they also
induce transformations of the (traditional) values. The (categorical or
homotopical) intuition here is that these axioms have computational content
because they witness isomorphisms rather than merely stating an extensional
equality.

The isomorphisms are extended to form a congruence relation by adding the
following constructors that witness equivalence and compatible closure:

\begin{table}[t]
\hrule\medskip
\begin{center} 
\Rule{}
{}
{\jdg{}{}{\idc : b \iso b}}
{}
\quad 
\Rule{}
{\jdg{}{}{c : b_1 \iso b_2}}
{\jdg{}{}{\symc{c} : b_2 \iso b_1}}
{}
\quad
\Rule{}
{\jdg{}{}{c_1 : b_1 \iso b_2} \quad c_2 : b_2 \iso b_3}
{\jdg{}{}{c_1 \fatsemi c_2 : b_1 \iso b_3}}
{}
\\ \bigskip
\Rule{}
{\jdg{}{}{c_1 : b_1 \iso b_2} \quad c_2 : b_3 \iso b_4}
{\jdg{}{}{c_1 \oplus c_2 : b_1 + b_3 \iso b_2 + b_4}}
{}
\quad 
\Rule{}
{\jdg{}{}{c_1 : b_1 \iso b_2} \quad c_2 : b_3 \iso b_4}
{\jdg{}{}{c_1 \otimes c_2 : b_1 * b_3 \iso b_2 * b_4}}
{}
\end{center}
\caption{Combinators\label{pi-combinators}}
\medskip\hrule
\end{table}

It is important to note that ``values'' and ``isomorphisms'' are completely
separate syntactic categories which do not intermix. The semantics of the
language come when these are made to interact at the ``top level'' via
\emph{application}: 
\[\begin{array}{lrcl}
\textit{top level term}, l &::=& c~v
\end{array}\]
